---
title: "ENM Day 2: Dismo"
output: 
  html_notebook:
    number_sections: true
    theme: cerulean
    toc: true
---
in dismo you just need a coordinates
long and lat only

Load dismo package
```{r libraries, echo=TRUE, warnings = FALSE, results="hide"}
library(raster)
library(dismo)
library(rgdal)
library(here)
```

Set paths to the files 
```{r}
# path to original layers
path2layers<-here("layers","worldclim","current")
# path to sp
path2sp<-here("layers","worldclim","species","dismo_rana.txt")
# path to training data
path2train<-here("results","dismo","rana_data.train.txt") # train
# path to test data
path2test<-here("results","dismo","rana_data.test.txt") #test data
# path to background data
path2backg<-here("results","dismo","rana_data.backgr.txt") # absense data

# path to bioclim model
path2bioclim<-here("results","dismo","bioclim","bc_rana")
# path to bioclim thresholded model))
path2bioclimtr<-here("results","dismo","bioclim","tr_bc_rana")
# path to domain model
path2domain<-here("results","dismo","domain","dm_rana")
# path to dismo thresholded model
path2domaintr<-here("results","dismo","domain","tr_dm_rana")


# path to future 2020 layers
path2020<-here("layers","worldclim","future","Iberia_wcB2a_10_hadcm3_2020")
# path to future 2050 layers
path2050<-here("layers","worldclim","future","Iberia_wcB2a_10_hadcm3_2050")
# path to future 2080 layers
path2080<-here("layers","worldclim","future","Iberia_wcB2a_10_hadcm3_2080")

# path to projected 2020 bioclim model
path2bioclim2020<-here("results","dismo","bioclim","bc_rana_2020")
# path to projected 2050 bioclim model
path2bioclim2050<-here("results","dismo","bioclim","bc_rana_2050")
# path to projected 2080 bioclim model
path2bioclim2080<-here("results","dismo","bioclim","bc_rana_2080")

# path to projected 2020 domain model
path2domain2020<-here("results","dismo","domain","dm_rana_2020")
# path to projected 2050 domain model
path2domain2050<-here("results","dismo","domain","dm_rana_2050")
# path to projected 2080 domain model
path2domain2080<-here("results","dismo","domain","dm_rana_2080")
```

# PREPARING DATA

Load environmental variables
```{r}
layers<-list.files(path=path2layers,pattern='asc',full.names=TRUE)
layers
```
Make the stack 
```{r}
egv<-stack(layers)
# Check egv structure
print(egv)
```

Descriptive statistics for environmental layers
```{r}
summary(egv)
```

Select variables to include in the model


Envelope algorythms are not affected by correlation between envir variables.so not nessasarily to select var with lower correlation
```{r}
names(egv)
#egv<-egv[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
```
Check egv structure
```{r}
print(egv)
```

Plot environmental variables on geographical space
```{r}
plot(egv,nc = 5)
```

Load species records
```{r}
data.sp<-read.csv(path2sp,sep=",",header=T)
# Check the structure of the species dataframe
head(data.sp)
```
Plot species records on geographical space
```{r}
plot(data.sp)
```

## Split species records in training and testing data

Using `sample()` function. 0.3 corresponds to testing proportion. 
30 % is test data, 70% are training data
```{r}
indexes = sample(1:nrow(data.sp),size = 0.3*nrow(data.sp))
## Split data in training and testing
data.train = data.sp[-indexes,]
data.test = data.sp[indexes,]
```

Check training dataframe
```{r}
head(data.train)
str(data.train)
```

Check testing dataframe
```{r}
head(data.test)
str(data.test)
```


## Extract background data

Create a cloud of random points
this is absenses

if we dont put `data.train,excludep= T` we generate the background
```{r}
abs<-as.data.frame(randomPoints(egv,500, data.train,excludep= T))
# to ensure that absenses are not close to presenses: data.train,excludep= T
#Check background dataframe
head(abs)
str(abs)
```


Export training, test, and background data
```{r}
write.table(data.train, file = path2train, sep='')
write.table(data.test, file = path2test, sep='')
write.table(abs, file = path2backg, sep='')
```

## Plot training, test, and absense data on geographical space
```{r}
par(mfrow=c(1,3))
plot(data.train,main="training data for Rana")
plot(data.test,main="test data for Rana")
plot(abs,main="absense data for Rana")

# Reset par function
par(mfrow=c(1,1))

```

# BIOCLIM Model

Calculate model
put env and test data
```{r}
bc<-bioclim(egv,data.train)
bc
```

## Plot presence and absence (background) data

1. Colinearity in the environmental data

```{r}
# most likely Pearson
pairs(bc)
```

2. Plot the model in the environmental space
    a: variable in X axis
    b: variable in Y axis
    p: percentile threshold
1 is 1st var, 2 is 2nd variable: they are stored alphabetically
red-excluded
green included
```{r}
plot(bc,a=1,b=2,p=1)
plot(bc,a=1,b=2,p=0.95)
plot(bc,a=1,b=2,p=0.90)
plot(bc,a=1,b=2,p=0.85)
plot(bc,a=1,b=2,p=0.80)
plot(bc,a=1,b=2,p=0.75)
plot(bc,a=1,b=2,p=0.70)
plot(bc,a=1,b=2,p=0.65)
plot(bc,a=1,b=2,p=0.60)
plot(bc,a=1,b=2,p=0.50)
```
    
3. Project the model
```{r}
p.bc<-predict(bc,egv)
p.bc
```

4. Plot the model in the geographical space

habitat suitability index: 1- perfect. des not have to do for presense of the species. by default "p = 95" not a p-value
```{r}
plot(p.bc)
```
5. Plot variable response curves
```{r}
response(bc)
```

##  Export Bioclim model

check raster formats available to export the results ?writeFormats. "HFA" is .img
```{r}

writeRaster(p.bc,path2bioclim,format="HFA",overwrite=TRUE)
```

## Evaluate the model

```{r}
e.bc<-dismo::evaluate(data.test,abs,bc,egv)
e.bc
```
  - 251 prsense, 500 absenses
  - max TPR+TNR is a Threshold

Check the evaluation results
```{r}
str(e.bc)
```

## Boxplot of presence and absence suitable values

blue are absences, red are presences
```{r}
boxplot(e.bc,col=c('blue','red'), notch=FALSE)
```

## Density plot of presence and absence suitable values

how well destingish the absenses agains the presenses- overlap quite a lot, so that means that there are absenses with similar env var as presenses

blue are absences, red are presences
```{r}
density(e.bc)
```

## Evaluation plots

```{r}
par(mfrow=c(3,3))
plot(e.bc,'ROC')
plot(e.bc,'CCR')
plot(e.bc,'TPR')
plot(e.bc,'TNR')
plot(e.bc,'FPR')
plot(e.bc,'FNR')
plot(e.bc,'PPP')
plot(e.bc,'NPP')
plot(e.bc,'kappa')
#Reset par function
par(mfrow=c(1,1))
```

## Obtain the threshold

- kappa: the threshold at which kappa is highest ("max kappa")
- spec_sens: the threshold at which the sum of the sensitivity (true positive rate) and specificity (true negative rate) is highest
- no_omission: the highest threshold at which there is no omission
- prevalence: modeled prevalence is closest to observed prevalence
- equal_sens_spec: equal sensitivity and specificity
- sensitivity: fixed (specified) sensitivity
```{r}
tr.b<-threshold(e.bc,'spec_sens')
# Get the value of the threshold
tr.b
```
presense only data has very small threshhold 

## Calculate the thresholded model

to select all pixels that pass a treshold 
```{r}
tr.bc<- p.bc > tr.b
tr.bc
```

Plot the p/a model in the geographical space
```{r}
plot(tr.bc)
```

Export p/a model
```{r}
writeRaster(tr.bc,path2bioclimtr,format="HFA",overwrite=TRUE)
```

## Plot raw model and p/a models (threshold model)

```{r}
par(mfrow=c(1,2))
plot(p.bc,main='Bioclim, raw values')
points(data.train,pch='+')
plot(tr.bc,main='presence/absence')
points(data.train,pch='+')
# Reset par function
par(mfrow=c(1,1))
```
to be very conservative choose teh highest threshold 

# DOMAIN Model

Calculate the model
```{r}
dm<-domain(egv,data.train)
dm
```

## Plot presence and absence (background) data

Colinearity in the environmental data
```{r}
pairs(dm)
```

Project the model
```{r}
p.dm<-predict(egv,dm)
p.dm
```
Plot the model in the geographical space
```{r}
plot(p.dm)
```
Plot variable response curves
```{r}
response(dm)
```

## Export Domain model

```{r}
writeRaster(p.dm,path2domain,format="HFA",overwrite=TRUE)
```

## Evaluate the model

```{r}
e.dm<-dismo::evaluate(data.test,abs,dm,egv)
e.dm
```

Check the evaluation results
```{r}
str(e.dm)
```

Boxplot of presence and absence suitable values
blue are absences, red are presences
```{r}
boxplot(e.dm,col=c('blue','red'))
```

Density plot of presence and absence suitable values
 blue are absences, red are presences
```{r}
density(e.dm)
```

## Evaluation plots

1. ROC
```{r}
par(mfrow=c(3,3))
plot(e.dm,'ROC')
plot(e.dm,'CCR') #CCR: Correct classification rate
plot(e.dm,'TPR') #TPR: True positive rate
plot(e.dm,'TNR') #TNR: True negative rate
plot(e.dm,'FPR') #FPR: False positive rate
plot(e.dm,'FNR') #FNR: False negative rate
plot(e.dm,'PPP') #PPP: Positive predictive power
plot(e.dm,'NPP') #NPP: Negative predictive power
plot(e.dm,'kappa') # kappa: Cohen’s kappa
#Reset par function
par(mfrow=c(1,1))
```

## Obtain the threshold

- kappa: the threshold at which kappa is highest ("max kappa")
- spec_sens: the threshold at which the sum of the sensitivity (true positive rate) and specificity (true negative rate) is highest
- no_omission: the highest threshold at which there is no omission
- prevalence: modeled prevalence is closest to observed prevalence
- equal_sens_spec: equal sensitivity and specificity
- sensitivty: fixed (specified) sensitivity
```{r}
tr.d<-threshold(e.dm,'spec_sens')
# Get the value of the threshold
tr.d
```
threshhold is high probably because domain is not very well implemented in DISMO

## Calculate the thresholded model

```{r}
tr.dm<-p.dm>tr.d
tr.dm
```

## Plot the p/a model in the geographical space

```{r}
plot(tr.dm)
```

Export p/a model
```{r}
writeRaster(tr.dm,path2domaintr,format="HFA",overwrite=TRUE)
```

## Plot raw and p/a models

```{r}
par(mfrow=c(1,2))
plot(p.dm,main='Domain, raw values')
points(data.train,pch='+')
plot(tr.dm,main='presence/absence')
points(data.train,pch='+')
# Reset par function
par(mfrow=c(1,1))
```

# Plot Comparision Bioclim and Domain

key difference between teh models:
study area for var 1 and var 2 ( x vs y) 

- DISMO draws a rectangle over all the points 
- DOMAIN draws more fitted polygons


```{r}
par(mfrow=c(1,2))
plot(tr.bc,main='Bioclim')
plot(tr.dm,main='Domain')
# Reset par function
par(mfrow=c(1,1))
```

# ENSEMBLE MODELS

## Mean raw model
```{r}
avg<-((p.bc+p.dm)/2)

plot(avg)
```

## Mean p/a model

mean of the threshold models 
```{r}
avg2<-((tr.dm+tr.bc)/2)
plot(avg2)
```

## Mean model weighted by AUC values
```{r}
avg3<-(((p.bc*e.bc@auc)+(p.dm*e.dm@auc))/2)
plot(avg3)
```

Plot mean models
```{r}
par(mfrow=c(1,2))
plot(avg)
plot(avg2)
```

Difference between p/a models
```{r}
par(mfrow=c(1,1))
diff.m<-abs(tr.dm-tr.bc)
plot(diff.m)
```
difference shows areas of disagreement between the models 

# PROJECTING MODELS TO OTHER SCENARIOS

## 2020
Load environmental variables for 2020
```{r}
layers2020<-list.files(path=path2020,pattern='asc',full.names=TRUE)
```

Make the stack
```{r}
egv2020<-stack(layers2020)
# Check egv structure
print(egv2020)
```

Descriptive statistics for environmental layers
```{r}
summary(egv2020)
```

Select variables to include in the model
```{r}
names(egv2020)
#egv2020<-egv2020[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check egv structure
print(egv2020)
```

Plot environmental variables on geographical space
```{r}
plot(egv2020)
```

## 2050
Load environmental variables for 2050
```{r}
layers2050<-list.files(path=path2050,pattern='asc',full.names=TRUE)
# Make the stack
egv2050<-stack(layers2050)
# Check egv structure
print(egv2050)
```

 Descriptive statistics for environmental layers
```{r}
summary(egv2050)
```
 
Select variables to include in the model
```{r}
names(egv2050)
#egv2050<-egv2050[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check egv structure
print(egv2050)
```

Plot environmental variables on geographical space
```{r}
plot(egv2050)
```

## 2080
Load environmental variables for 2080
```{r}
layers2080<-list.files(path=path2080,pattern='asc',full.names=TRUE)
# Make the stack
egv2080<-stack(layers2080)
# Check egv structure
print(egv2080)
```

Descriptive statistics for environmental layers
```{r}
summary(egv2080)
```

Select variables to include in the model
```{r}
names(egv2080)
#egv2080<-egv2080[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check egv structure
print(egv2080)
```

Plot environmental variables on geographical space
```{r}
plot(egv2080)
```

# Project models to future scenarios

## BIOCLIM
### Project the model to 2020
```{r}
p.bc.20 = predict(egv2020,bc)
p.bc.20
```

# Plot the model in the geographical space
```{r}
plot(p.bc.20)
```

Export Bioclim 2020 model
```{r}
writeRaster(p.bc.20,path2bioclim2020,format="HFA",overwrite=TRUE)
```

### Project the model to 2050
```{r}
p.bc.50 = predict(egv2050,bc)
p.bc.50
```

Plot the model in the geographical space
```{r}
plot(p.bc.50)
```

Export Bioclim 2050 model
```{r}
writeRaster(p.bc.50,path2bioclim2050,format="HFA",overwrite=TRUE)
```

### Project the model to 2080
```{r}
p.bc.80 = predict(egv2080,bc)
p.bc.80
```

Plot the model in the geographical space
```{r}
plot(p.bc.80)
```
Export Bioclim 2080 model
```{r}
writeRaster(p.bc.80,path2bioclim2080,format="HFA",overwrite=TRUE)
```

### Plot Bioclim current and future models
```{r}
par(mfrow=c(2,2))
plot(p.bc,main='Bioclim current')
plot(p.bc.20,main='Bioclim 2020')
plot(p.bc.50,main='Bioclim 2050')
plot(p.bc.80,main='Bioclim 2080')
# Reset par function
par(mfrow=c(1,1))
```

## DOMAIN

### Project the model to 2020
```{r}
p.dm.20 = predict(egv2020, dm)
p.dm.20
```


 Plot the model in the geographical space
```{r}
plot(p.dm.20)
```

Export domain 2020 model
```{r}
writeRaster(p.dm.20,path2domain2020,format="HFA",overwrite=TRUE)
```


### Project the model to 2050
```{r}
p.dm.50 = predict(egv2050,dm)
p.dm.50
```

# Plot the model in the geographical space
```{r}
plot(p.dm.50)
```

# Export domain 2050 model
```{r}
writeRaster(p.dm.50,path2domain2050,format="HFA",overwrite=TRUE)
```


### Project the model to 2080

```{r}
p.dm.80 = predict(egv2080,dm)
p.dm.80
```

Plot the model in the geographical space
```{r}
plot(p.dm.80)
```

Export domain 2080 model
```{r}
writeRaster(p.dm.80,path2domain2080,format="HFA",overwrite=TRUE)
```


Plot domain current and future models
```{r}
par(mfrow=c(2,2))
plot(p.dm,main='Domain current')
plot(p.dm.20,main='Domain 2020')
plot(p.dm.50,main='Domain 2050')
plot(p.dm.80,main='Domain 2080')
#Reset par function
par(mfrow=c(1,1))
```

## Plot bioclim and domain current and future models
```{r}
par(mfrow=c(2,4))
plot(p.bc,main='Bioclim current')
plot(p.bc.20,main='Bioclim 2020')
plot(p.bc.50,main='Bioclim 2050')
plot(p.bc.80,main='Bioclim 2080')
plot(p.dm,main='Domain current')
plot(p.dm.20,main='Domain 2020')
plot(p.dm.50,main='Domain 2050')
plot(p.dm.80,main='Domain 2080')
#Reset par function
par(mfrow=c(1,1))
```

## Threshold all models
```{r}
tr.bc.20<-p.bc.20>tr.b
tr.bc.50<-p.bc.50>tr.b
tr.bc.80<-p.bc.80>tr.b
tr.dm.20<-p.dm.20>tr.d
tr.dm.50<-p.dm.50>tr.d
tr.dm.80<-p.dm.80>tr.d
```

Plot thresholded bioclim and domain current and future models
```{r}
par(mfrow=c(2,4))
plot(tr.bc,main='Bioclim current')
plot(tr.bc.20,main='Bioclim 2020')
plot(tr.bc.50,main='Bioclim 2050')
plot(tr.bc.80,main='Bioclim 2080')
points(data.train,pch='+')
plot(tr.dm,main='Domain current')
plot(tr.dm.20,main='Domain 2020')
plot(tr.dm.50,main='Domain 2050')
plot(tr.dm.80,main='Domain 2080')
points(data.train,pch='+')
#Reset par function
par(mfrow=c(1,1))
```

