---
title: 'ENM Day 3: Simple Species Modeling Biomod'
output:
  word_document:
    toc: yes
  html_notebook:
    number_sections: yes
    theme: cerulean
    toc: yes
---
# Biomod2 without pseudo-absences

load the library
```{r, results = "hide"}
library(biomod2)
library(here)
library(ggplot2)
library(dplyr)
library(readr)
```

Biomod will automatically create directories to put files in.
```{r}
# path to species
path2sp<-here("layers","worldclim","species","biomod_rana.csv")
# path to layers
path2layers<-here("layers","worldclim","current")
# path to working directory
path2work<-here()
# path to Biomod models
path2models<-here("Rana","models","allmodels")
# path to current models
path2current<-here("Rana","proj_current")
# path to future 2020 laters
path2020<-here("layers","worldclim","future","Iberia_wcB2a_10_hadcm3_2020")
# path to future 2020 models
path2models2020<-here("Rana","proj_future2020")
# path to future 2050 laters
path2050<-here("layers","worldclim","future","Iberia_wcB2a_10_hadcm3_2050")
# path to future 2050 models
path2models2050<-here("Rana","proj_future2050")
# path to future 2080 laters
path2080<-here("layers","worldclim","future","Iberia_wcB2a_10_hadcm3_2080")
# path to future 2080 models
path2models2080<-here("Rana","proj_future2080")
# path to pseudo-absence models
path2psmodels<-here("Rana","models","all_ps2_models")
# path to species without absences
path2sp2<-here("layers","worldclim","species","biomod_rana2.csv")
```

##  loading_data

load our species data
```{r}
DataSpecies <- read.csv(path2sp,sep=',')
# Check species data
head(DataSpecies)
```

plot species data
```{r}
DataSpecies %>% 
  ggplot() +geom_point(aes(x=X_WGS84,y=Y_WGS84,colour = factor(Rana)))+
  scale_color_manual(values = c("red", "green"))+
  theme_bw()
```

biomid require to idu=icate couple of things:
- the name of studied species
```{r}
myRespName <- 'Rana' # match teh one in dataframe 
```

- the presence/absences data for our species 
```{r}
myResp <- as.numeric(DataSpecies[,myRespName])
myResp
```

- The XY coordinates of species data
```{r}
myRespXY <- DataSpecies[,c("X_WGS84","Y_WGS84")]
# check coordinates
str(myRespXY)
head(myRespXY)
```

## Load Environmental variables

```{r}
layers<-stack(list.files(path=path2layers,pattern='asc',full.names=TRUE ))
# Check layers structure
print(layers)
```

Plot layers
```{r}
plot(layers)
```
 
Descriptive statistics for environmental layers
```{r}
summary(layers)
```

## Select variables to include in the model 

(low correlation) for backround and presense/absense algotrythms is a must to select low correlated 
```{r}
names(layers)
myExpl<-layers[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check layers structure
print(myExpl)
```

Plot environmental variables on geographical space
```{r}
plot(myExpl)
```


##  2: formating_data

Check help to know how to define pseudo-absences
?BIOMOD_FormatingData

we dont need to define pseudoabsenses: we have absenses
to define pseudoabsenses
```{r}
myBiomodData <- BIOMOD_FormatingData(resp.var = myResp,
                                     expl.var = myExpl,
                                     resp.xy = myRespXY,
                                     resp.name = myRespName #name of models usually species name (biomod create a folder with sp name)
#                                     PA.nb.rep = 1, # number of pseudo-absence datasets
#                                     PA:nb.absences = 200, # number of pseudo-absence records
#                                     PA.strategy = 'random' # pseudo-absences are selected randomly
                                    )
# print_formating_data
myBiomodData
```
in regressions: same as prsense data ( better a bit more: 50 p, 60 absenses)
RF you need a lot more 
this ex model everything 


plot_formating_data
```{r}
plot(myBiomodData)
```

## modeling_options

Defining Models Options using default options
Consult help `?BIOMOD_ModelingOptions`
```{r}
myBiomodOption <- BIOMOD_ModelingOptions()
```

## modeling

Computing the models 
See help
`?BIOMOD_Modeling`
models = c('GLM','GBM','GAM','CTA','ANN','SRE','FDA','MARS','RF')
```{r}
myBiomodModelOut <- BIOMOD_Modeling( 
                      myBiomodData, # data including species records, species name, and variables
                      models = c('GLM','GAM'), # modelling methods
                      models.options = myBiomodOption, # options for modelling (use default parameters)
                      NbRunEval=1, # number of Evaluation run. 10 is better
                      DataSplit=70, # % of data used as training
                      VarImport=3, # number of permutation to estimate var importance
                      models.eval.meth = c('TSS','ROC','KAPPA'), # evaluation metrics
                      SaveObj = TRUE, # keep all results and outputs on hard drive
                      rescal.all.models = TRUE, # all models scaled with a binomial GLM: the range of the scale is 0-1000.( all prob or hab suitability)
                      do.full.models = TRUE, # models calibrated and evaluated with the whole dataset
                      modeling.id = "allmodels") # ID (=name) of modeling procedure
```
ignore the warning 
files created on hard drive
```{r}
list.files(path2models)
```
Output will have for each model (GAM and GLM) Run1 and FUll all points. usually better than the runs as it have all data


modeling_summary
```{r}
myBiomodModelOut 
```
failed: glm=> did nlt converge 
if you have failed: need more data 

biomod usually do not compare algorythms, rather you just want to model species

## modeling evaluations

get all models evaluation
```{r}
myBiomodModelEval <- get_evaluations(myBiomodModelOut)
myBiomodModelEval

```
cutoff max value for probability bn 0-1000. 


 print the dimnames of this object
```{r}
 
dimnames(myBiomodModelEval)
```

print the TSS scores of GLM & GAM
```{r}
myBiomodModelEval["TSS","Testing.data","GAM",,]
```

print the ROC scores of all selected models
```{r}
myBiomodModelEval["ROC","Testing.data",,,]
```

print variable importances
```{r}
get_variables_importance(myBiomodModelOut)
```

save models evaluation scores and variables importance on hard drive
```{r}
capture.output(get_evaluations(myBiomodModelOut),
               file=file.path(myRespName, 
                              paste(myRespName,"_formal_models_evaluation.txt", sep="")))

capture.output(get_variables_importance(myBiomodModelOut),
               file=file.path(myRespName, 
                              paste(myRespName,"_formal_models_variables_importance.txt", sep="")))   
```



## projection of current models

projection over the study area under current conditions
```{r}
myBiomodProj <- BIOMOD_Projection(
  modeling.output = myBiomodModelOut, # modelling results
  new.env = myExpl, # environmental variables (here, current climate variables)
  proj.name = 'current', # name of the projections
  selected.models = 'all', # models to be projected
  do.stack = FALSE,# True defaul. false 1 file per model
  binary.meth = 'TSS', # a vector of a subset of models evaluation method computed before
  build.clamping.mask = T, # if TRUE, a clamping mask will be saved on hard drive different
  output.format = '.grd') # the format of the GIS files: RData, .grd ( saves names)  or .img (GIS format, cant see names of layers)

# summary of created object
myBiomodProj
```
output files: you dont care aux.xml,
TSSbin: trshold model 
clamping map gives you the area outside teh range of teh model 

files created on hard drive
```{r}
list.files(path2current)
```

## Ensemble modeling

`?BIOMOD_EnsembleModeling`

decide which model to include 
```{r}
myBiomodEM <- BIOMOD_EnsembleModeling( 
                     modeling.output = myBiomodModelOut, # model results
                     chosen.models = 'all', # models to be included when ensembling
                     em.by='all', # flag defining the way the models will be combined to build the ensemble models: 'PA_dataset+repet' (default), 'PA_dataset+algo', 'PA_dataset', 'algo', 'all'
                     # See the vignette: http://127.0.0.1:9144/help/library/biomod2/doc/EnsembleModelingAssembly.pdf
                     eval.metric = c('TSS'), # evaluation metric used to build ensemble models
                     eval.metric.quality.threshold = c(0.7), # If not NULL, the minimum scores below which models will be excluded of the ensemble-models building
                     prob.mean = T, # estimate the mean probabilities across predictions
                     prob.cv = T, # estimate the coefficient of variation across predictions
                     prob.ci = T, # estimate the confidence interval around the prob.mean
                     prob.ci.alpha = 0.05, # significance level for estimating the confidence interval. Default = 0.05
                     prob.median = T, # estimate the mediane of probabilities
                     committee.averaging = T, # estimate the committee averaging (species richness map) across predictions
                     # The committee averaging score is then the average of binary predictions
                     prob.mean.weight = T, # estimate the weighted sum of probabilities
                     prob.mean.weight.decay = 'proportional' ) # define the relative importance of the weights

# print summary                     
myBiomodEM
```

-weighted mean is a better as you weight it based on teh quality of teh model
-mean produce more smooth maps


get evaluation scores
```{r}
get_evaluations(myBiomodEM)
```

##  EnsembleForecasting_current
```{r}
myBiomodEF <- BIOMOD_EnsembleForecasting( 
  EM.output = myBiomodEM,
  projection.output = myBiomodProj,
  output.format = '.img')

# print summary
myBiomodEF
```
 files created on hard drive
```{r}
list.files(path2current)
```

##  Projection to future conditions

### 2020

load environmental variables for the future
```{r}
layers.2020<-stack(list.files(path=path2020,pattern='asc',full.names=TRUE))
# Check layers structure
print(layers.2020)
# Descriptive statistics for environmental layers
summary(layers.2020)
```


Select variables to include in the model
```{r}
names(layers.2020)
myExpl2020<-layers.2020[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check layers structure
print(myExpl2020)
```

Plot environmental variables on geographical space
```{r}
plot(myExpl2020)
```


```{r}
myBiomodProj2020 <- BIOMOD_Projection(
                              modeling.output = myBiomodModelOut,
                              new.env = myExpl2020,
                              proj.name = 'future2020',
                              selected.models = 'all',
                              binary.meth = 'TSS',
                              build.clamping.mask = T,
                              output.format = '.img')

# print summary
myBiomodProj2020
```

files created on hard drive
```{r}
list.files(path2models2020)
```



### 2050

load environmental variables for the future. 
```{r}
layers.2050<-stack(list.files(path=path2050,pattern='asc',full.names=TRUE))
# Check layers structure
print(layers.2050)
```
Descriptive statistics for environmental layers
```{r}
summary(layers.2050)
```


Select variables to include in the model
```{r}
names(layers.2050)
myExpl2050<-layers.2050[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check layers structure
print(myExpl2050)
```

Plot environmental variables on geographical space
```{r}
plot(myExpl2050)
```


```{r}
myBiomodProj2050 <- BIOMOD_Projection(
  modeling.output = myBiomodModelOut,
  new.env = myExpl2050,
  proj.name = 'future2050',
  selected.models = 'all',
  binary.meth = 'TSS',
  build.clamping.mask = T,
  output.format = '.img')

# print summary
myBiomodProj2050
```

files created on hard drive
```{r}
list.files(path2models2050)
```

### 2080

load environmental variables for the future. 
```{r}

layers.2080<-stack(list.files(path=path2080,pattern='asc',full.names=TRUE))
# Check layers structure
print(layers.2080)
```

Descriptive statistics for environmental layers
```{r}
summary(layers.2080)
```


Select variables to include in the model
```{r}

names(layers.2080)
myExpl2080<-layers.2080[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
```

 Check layers structure
```{r}
print(myExpl2080)
```


## Plot environmental variables on geographical space
```{r}
plot(myExpl2080)
```


```{r}
myBiomodProj2080 <- BIOMOD_Projection(
  modeling.output = myBiomodModelOut,
  new.env = myExpl2080,
  proj.name = 'future2080',
  selected.models = 'all',
  binary.meth = 'TSS',
  build.clamping.mask = T,
  output.format = '.img')

# print summary
myBiomodProj2080
```

files created on hard drive
```{r}
list.files(path2models2080)
```



## Plot current and future models

```{r}
#current
plot(myBiomodProj)
#future
plot(myBiomodProj2020)
plot(myBiomodProj2050)
plot(myBiomodProj2080)
# pick a certain model
plot(myBiomodProj,str.grep='GLM')
plot(myBiomodProj2020,str.grep='RUN1_GLM')
plot(myBiomodProj2050,str.grep='GLM')
plot(myBiomodProj2080,str.grep='Full_GAM')
```

# Biomod2 with pseudo-absences

## formating_data

Check help to know how to define pseudo-absences
?BIOMOD_FormatingData

check stability of teh model based onnumber of PA provided

if sat has A, it will be ignored
```{r}

myBiomodData <- BIOMOD_FormatingData(resp.var = myResp,
                                     expl.var = myExpl,
                                     resp.xy = myRespXY,
                                     resp.name = myRespName,
                                     PA.nb.rep = 1, # number of pseudo-absence datasets. better put 10
                                     PA.nb.absences = 800, # number of pseudo-absence records
                                     PA.strategy = 'disk', # pseudo-absences are selected randomly. disk: distance for presenses
                                     PA.dist.min = 0.08333334, # in degress,  minimal distance to presences : no absenses in teh same pixel where you have prsense
                                     PA.dist.max = NULL # maximal distance to presences 
                                     )

```

print_formating_data
```{r}
myBiomodData
```



## plot_formating_data
```{r}
plot(myBiomodData)

```

## Modeling_options

Defining Models Options using default options
Consult help
?BIOMOD_ModelingOptions
```{r}
myBiomodOption <- BIOMOD_ModelingOptions()
```


## Modeling

Computing the models 
See help
?BIOMOD_Modeling
```{r}
myBiomodModelOut <- BIOMOD_Modeling( 
  myBiomodData, # data including species records, species name, and variables
  models = c('RF'), # modelling methods
  models.options = myBiomodOption, # options for modelling
  NbRunEval=3, # number of Evaluation run
  DataSplit=70, # % of data used as training
  VarImport=3, # number of permutation to estimate var importance
  models.eval.meth = c('TSS','ROC'), # evaluation metrics
  SaveObj = TRUE, # keep all results and outputs on hard drive
  rescal.all.models = TRUE, # all models scaled with a binomial GLM: the range of the scale is 0-1000
  do.full.models = FALSE, # models calibrated and evaluated with the whole dataset. TRUE if you want a model with all teh models
  modeling.id = "all_ps2_models") # ID (=name) of modeling procedure

```

files created on hard drive
```{r}
list.files(path2psmodels)
```

modeling_summary
```{r}
myBiomodModelOut 
```


## modeling evaluations

get all models evaluation

check AUC between training and test data: should be similar.
```{r}

myBiomodModelEval <- get_evaluations(myBiomodModelOut)
myBiomodModelEval
```


print the dimnames of this object

```{r}
dimnames(myBiomodModelEval)
```

print the ROC scores of all selected models
```{r}
myBiomodModelEval["ROC","Testing.data",,,]
```

print variable importances
```{r}
get_variables_importance(myBiomodModelOut)
```

save models evaluation scores and variables importance on hard drive
```{r}
capture.output(get_evaluations(myBiomodModelOut),
               file=file.path(myRespName, 
                              paste(myRespName,"_formal_models_evaluation.txt", sep="")))

capture.output(get_variables_importance(myBiomodModelOut),
               file=file.path(myRespName, 
                              paste(myRespName,"_formal_models_variables_importance.txt", sep="")))  
```


## projection of current models

projection over the study area under current conditions
```{r}
myBiomodProj <- BIOMOD_Projection(
  modeling.output = myBiomodModelOut, # modelling results
  new.env = myExpl, # environmental variables (here, current climate variables)
  proj.name = 'current', # name of the projections
  selected.models = 'all', # models to be projected
  binary.meth = 'ROC', # a vector of a subset of models evaluation method computed before
  build.clamping.mask = T, # if TRUE, a clamping mask will be saved on hard drive different
  output.format = '.grd') # the format of the GIS files: .RData, .grd, or .img

```


summary of created object
```{r}
myBiomodProj
```


files created on hard drive
```{r}
list.files(path2current)
```

## Plot current models

```{r}
#current
plot(myBiomodProj)
```
## plot 2D response plots

```{r}
myRF <- BIOMOD_LoadModels(myBiomodModelOut, models='RF')
```

response curve shows what it prefers
```{r}


myRespPlot2D <- response.plot2(models  = myRF,
                               Data = get_formal_data(myBiomodModelOut,'expl.var'), 
                               show.variables= get_formal_data(myBiomodModelOut,'expl.var.names'),
                               do.bivariate = FALSE,
                               fixed.var.metric = 'median',
                               col = c("blue", "red"),
                               legend = TRUE,
                               data_species = get_formal_data(myBiomodModelOut,'resp.var'))




```

- you want to see smth like bio18: optimal levels
- if only one limit (bio16) means there is no limit in a given area

## plot 3D response plots

here only for a lone model
```{r, fig.width=20, fig.length=20}
myRespPlot3D <- response.plot2(models  = myRF[1],
                               Data = get_formal_data(myBiomodModelOut,'expl.var'), 
                               show.variables= get_formal_data(myBiomodModelOut,'expl.var.names'),
                               do.bivariate = TRUE,
                               fixed.var.metric = 'median',
                               data_species = get_formal_data(myBiomodModelOut,'resp.var'),
                               display_title=FALSE)
```

