---
title: "ENM Day 4: Simple Species Modeling MAXENT"
output: 
  html_notebook:
    number_sections: true
    theme: cerulean
    toc: true
---

#####################################################
############### MAXENT WITH DISMO ###################
#####################################################
MAXENT wont work if Folders are named with a space !!
# Load libraries
```{r}
library(raster)
library(rgdal)
library(maptools)
library(here)
# Load dismo package
library(dismo)
library(rJava)
```


```{r}

# path to species
path2sp<-here("layers","worldclim","species","dismo_rana.txt")
# path to layers
path2layers<-here("layers","worldclim","current")
# path to data train database
path2train<-here("results","dismo","rana_data.train.txt")
# path to data test database
path2test<-here("results","dismo","rana_data.test.txt")
# path to data background database
path2backg<-here("results","dismo","rana_data.backgr.txt")
# path to maxent jar file in dismo
path2jar<-here("results","dismo","maxent")
# path to maxent tif model
path2tif<-here("results","dismo","maxent","maxent_prediction_rana.tif")
# path to maxent model
path2maxent<-here("results","dismo","maxent","me.rana")
# path to maxent thresholded model
path2maxenttr<-here("results","dismo","maxent","tr_me_rana")
# path to future 2020 laters
path2020<-here("layers","worldclim","future","Iberia_wcB2a_10_hadcm3_2020")
# path to future 2020 models
path2models2020<-here("results","dismo","maxent","me_rana_2020")
# path to future 2050 laters
path2050<-here("layers","worldclim","future","Iberia_wcB2a_10_hadcm3_2050")
# path to future 2050 models
path2models2050<-here("results","dismo","maxent","me_rana_2050")
# path to future 2080 laters
path2080<-here("layers","worldclim","future","Iberia_wcB2a_10_hadcm3_2080")
# path to future 2080 models
path2models2080<-here("results","dismo","maxent","me_rana_2080")

```


######################################################
# PREPARING DATA

# Load environmental variables
```{r}
layers<-list.files(path=path2layers,pattern='asc',full.names=TRUE)
# Make the stack
egv<-stack(layers)
# Check the stack structure
print(egv)
```


# Descriptive statistics for environmental layers
```{r}
summary(egv)
# Select variables to include in the model
names(egv)
egv<-egv[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check the stack structure
print(egv)
```


# Plot environmental variables on geographical space
```{r}
plot(egv)
```


# Load species records
```{r}
data.sp<-read.csv(path2sp,sep=",",header=T)
# Check the structure of the species dataframe
head(data.sp)
# Plot species records on geographical space
plot(data.sp)
```

# Split species records in training and testing data
## Using sample function
## 0.3 corresponds to testing proportion
```{r}
indexes = sample(1:nrow(data.sp),size=0.3*nrow(data.sp))
```


## Split data in training and testing
```{r}
data.train = data.sp[-indexes,]
data.test = data.sp[indexes,]
# Check training dataframe
head(data.train)
str(data.train)
# Check testing dataframe
head(data.test)
str(data.test)
```

# Extract background data
```{r}
# Create a cloud of random points
backgr<-as.data.frame(randomPoints(egv,500))
# Check background dataframe
head(backgr)
str(backgr)
```


# Export training, test, and background data
```{r}
write.table(data.train,file=path2train,sep='')
write.table(data.test,file=path2test,sep='')
write.table(backgr,file=path2backg,sep='')
```



######################################################
# MAXENT WITH DISMO

# only run if the maxent.jar file is available,in the right folder
# Copy maxent.jar to /home/neftali/R/x86_64-pc-linux-gnu-library/3.2/dismo/java
```{r}
jar<-paste(system.file(package="dismo"),"/java/maxent.jar",sep='')
```


# Calculate model with MAXENT parameters
# Choose parameters
# See maxent function in R help
# See maxent parameters in maxent help
```{r}
?dismo::maxent
me<-maxent(egv,data.train,args=c("randomseed","writeclampgrid","betamultiplier=1","responsecurves","jackknife","pictures","replicates=1"),removeDuplicates=TRUE,path=path2jar)

```

# Calculate model
# 'biome' is a categorical variable
#me<-maxent(egv,data.train,factors='biome',....)
# See the maxent results in a browser
```{r}
me
```

# Plot importance of each variable
```{r}
plot(me)
```

# Response curves
```{r}
response(me)
```


# Project the model
```{r}
p.me<-predict(me,egv,args=c("outputformat=logistic"),progress='text',filename=path2tif,overwrite=TRUE)
plot(p.me)
points(data.train)
```


# Export Maxent model
```{r}
writeRaster(p.me,path2maxent,format="HFA",overwrite=TRUE)
```


# Evaluate the model
```{r}
e.me<-dismo::evaluate(me,p=data.test,a=backgr,x=egv)
e.me
```
AUC value is different from html generated  because we using test dat anad in html we uses train data

# Check the evaluation results
```{r}
str(e.me)
```


# Boxplot of presence and absence suitable values
# blue are absences, red are presences
```{r}
boxplot(e.me,col=c('blue','red'))
```


# Density plot of presence and absence suitable values
# blue are absences, red are presences
```{r}
density(e.me)
```


# Evaluation plots
```{r}
par(mfrow=c(3,3))
# ROC
plot(e.me,'ROC')
# CCR: Correct classification rate
plot(e.me,'CCR')
# TPR: True positive rate
plot(e.me,'TPR')
# TNR: True negative rate
plot(e.me,'TNR')
# FPR: False positive rate
plot(e.me,'FPR')
# FNR: False negative rate
plot(e.me,'FNR')
# PPP: Positive predictive power
plot(e.me,'PPP')
# NPP: Negative predictive power
plot(e.me,'NPP')
# kappa: Cohen's kappa
plot(e.me,'kappa')
par(mfrow=c(1,1))
```



# Obtain the threshold
# kappa: the threshold at which kappa is highest ("max kappa")
# spec_sens: the threshold at which the sum of the sensitivity (true positive rate) and specificity (true negative rate) is highest
# no_omission: the highest threshold at which there is no omission
# prevalence: modeled prevalence is closest to observed prevalence
# equal_sens_spec: equal sensitivity and specificity
# sensitivity: fixed (specified) sensitivity
```{r}
tr<-threshold(e.me,'prevalence')
# Get the value of the threshold
tr
```



# Get the value of all thresholds
```{r}
tr<-threshold(e.me)
tr[,6]
```


# Calculate the thresholded model
```{r}
tr.me<-p.me>(tr[,6])
# Plot the p/a model in the geographical space
tr.me
```


```{r}
plot(tr.me)
```


# Export p/a model
```{r}
writeRaster(tr.me,path2maxenttr,format="HFA",overwrite=TRUE)
```



# Plot raw and p/a models
```{r}
par(mfrow=c(1,2))
plot(p.me,main='Maxent, raw values')
points(data.train,pch='+')
plot(tr.me,main='presence/absence')
points(data.test,pch='+')

# Reset par function
par(mfrow=c(1,1))
```




######################################################
# PROJECTING MODELS TO OTHER SCENARIOS
```{r}
# Load environmental variables for 2020
layers2020<-list.files(path=path2020,pattern='asc',full.names=TRUE)
# Make the stack
egv2020<-stack(layers2020)
# Check the stack structure
print(egv2020)
# Descriptive statistics for environmental layers
summary(egv2020)
# Select variables to include in the model
egv2020<-egv2020[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check the stack structure
print(egv2020)
# Plot environmental variables on geographical space
plot(egv2020)
```



# Load environmental variables for 2050
```{r}
layers2050<-list.files(path=path2050,pattern='asc',full.names=TRUE)
# Make the stack
egv2050<-stack(layers2050)
# Check the stack structure
print(egv2050)
# Descriptive statistics for environmental layers
summary(egv2050)
# Select variables to include in the model
egv2050<-egv2050[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check the stack structure
print(egv2050)
```



# Load environmental variables for 2080
```{r}
layers2080<-list.files(path=path2080,pattern='asc',full.names=TRUE)
# Make the stack
egv2080<-stack(layers2080)
# Check the stack structure
print(egv2080)
# Descriptive statistics for environmental layers
summary(egv2080)
# Select variables to include in the model
egv2080<-egv2080[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Check the stack structure
print(egv2080)
# Plot environmental variables on geographical space
plot(egv2080)
```



# Project the model to 2020
```{r}
p.me.20 = predict(egv2020,me)
p.me.20
# Plot the model in the geographical space

plot(p.me.20)
```


# Export Maxent model
```{r}
writeRaster(p.me.20,path2models2020,format="HFA",overwrite=TRUE)
```



# Project the model to 2050
```{r}
p.me.50 = predict(egv2050,me)
p.me.50
# Plot the model in the geographical space

plot(p.me.50)
```


# Export Maxent model
```{r}
writeRaster(p.me.50,path2models2050,format="HFA",overwrite=TRUE)
```



# Project the model to 2080
```{r}
p.me.80 = predict(egv2080,me)
p.me.80
# Plot the model in the geographical space
plot(p.me.80)
```


# Export Maxent model
```{r}
writeRaster(p.me.80,path2models2080,format="HFA",overwrite=TRUE)
```



# Plot Maxent current and future models
```{r}
par(mfrow=c(2,2))
plot(p.me,main='Maxent current')
plot(p.me.20,main='Maxent 2020')
plot(p.me.50,main='Maxent 2050')
plot(p.me.80,main='Maxent 2080')
# Reset par function
par(mfrow=c(1,1))

```

# MAXENT WITH BIOMOD2



# load the library
```{r, results= "hide}
library(biomod2)
library(here)
```
path to files and locations

```{r}
library(here)
# path to species
path2sp<-here( "layers" , "worldclim" , "species" , "biomod_rana.csv")
# path to layers
path2layers<-here( "layers" , "worldclim" , "current")
# path to maxent jar file
path2maxent<-here( "R" , "04_maxent" , "maxent" )
# path to working directory
path2work<-here( "results" , "biomod" , "maxent" )
# path to future 2020 laters
path2020<-here( "layers" , "worldclim" , "future" , "Iberia_wcB2a_10_hadcm3_2020")
# path to future 2020 models
path2models2020<-here("Rana" , "proj_future2020")
# path to future 2050 laters
path2050<-here( "layers" , "worldclim" , "future" , "Iberia_wcB2a_10_hadcm3_2050")
# path to future 2050 models
path2models2050<-here( "Rana" , "proj_future2050")
# path to future 2080 laters
path2080<-here( "layers" , "worldclim" , "future" , "Iberia_wcB2a_10_hadcm3_2080")
# path to future 2080 models
path2models2080<-here( "Rana" , "proj_future2080")

```


########################################
# load our species data
```{r}
DataSpecies <- read.csv(path2sp,sep=',')
# Check species data
head(DataSpecies)
str(DataSpecies)
```



# the name of studied species
```{r}
myRespName <- 'Rana'

# the presence/absences data for our species 
myResp <- as.numeric(DataSpecies[,myRespName])

# the XY coordinates of species data
myRespXY <- DataSpecies[,c("X_WGS84","Y_WGS84")]

```



# load the environmental raster layers
```{r}
layers<-stack(list.files(path=path2layers,pattern='asc',full.names=TRUE ))
# Descriptive statistics for environmental layers
summary(layers)
```


# Select variables to include in the model
```{r}
names(layers)
myExpl<-layers[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
# Plot environmental variables on geographical space
plot(myExpl)
```



# Formating data
```{r}
myBiomodData <- BIOMOD_FormatingData(
  resp.var = myResp, 
  expl.var = myExpl, 
  resp.xy = myRespXY, 
  resp.name = myRespName)

# print_formating_data
myBiomodData

```

#: plot_formating_data
```{r}
plot(myBiomodData)
```
number of point per latitude : grey stuff at teh edges


# Defining MAXENT Mododelling options 
# Consult help
?BIOMOD_ModelingOptions
```{r}
?BIOMOD_ModelingOptions
myBiomodOption <- BIOMOD_ModelingOptions(MAXENT.Phillips = list(path_to_maxent.jar=path2maxent, 
                         memory_allocated=1024,
                         maximumiterations = 500, 
                         visible = TRUE, # to view maxent running
                         linear = TRUE,
                         quadratic = TRUE, 
                         product = TRUE, # not recommended
                         threshold = TRUE, 
                         hinge = TRUE, 
                         lq2lqptthreshold = 80, 
                         l2lqthreshold = 10, 
                         hingethreshold = 15, 
                         beta_threshold = -1, 
                         beta_categorical = -1, 
                         beta_lqp = -1, 
                         beta_hinge = -1, 
                         betamultiplier = 1,# changed
                         defaultprevalence = 0.5))

```


# Computing the models 
# See help
?BIOMOD_Modeling
```{r}
myBiomodModelOut <- BIOMOD_Modeling( myBiomodData, 
                                     models =  'MAXENT.Phillips',
                                     models.options = myBiomodOption, 
                                     NbRunEval=1, # number of Evaluation run
                                     DataSplit=70, # % of data used as training
                                     VarImport=3, # number of permutation to estimate var importance
                                     models.eval.meth = 'ROC', # evaluation metrics
                                     SaveObj = TRUE,# keep all results and outputs on hard drive
                                     rescal.all.models = FALSE, # all models scaled with a binomial GLM
                                     modeling.id = "maxent") # ID (=name) of modeling procedure


# modeling_summary
myBiomodModelOut
```



# get all models evaluation 
```{r}
myBiomodModelEval<-get_evaluations(myBiomodModelOut) 
myBiomodModelEval
```



# print the dimnames of this object
```{r}
dimnames(myBiomodModelEval)
```



# print the ROC scores of all selected models
```{r}
myBiomodModelEval["ROC","Testing.data",,,]
```



# print variable importances
```{r}
get_variables_importance(myBiomodModelOut)
```



# save models evaluation scores and variables importance on hard drive
```{r}
capture.output(get_evaluations(myBiomodModelOut),
               file=file.path(myRespName, 
                              paste(myRespName,"_formal_models_evaluation.txt", sep="")))

capture.output(get_variables_importance(myBiomodModelOut),
               file=file.path(myRespName, 
                              paste(myRespName,"_formal_models_variables_importance.txt", sep="")))  

```


# Project models over studied area 
```{r}
myBiomodProj <- BIOMOD_Projection(
  modeling.output = myBiomodModelOut, # modelling results
  new.env = myExpl, # environmental variables (here, current climate variables)
  proj.name = 'current', # name of the projections
  selected.models = 'all', # models to be projected
  binary.meth = 'ROC', # a vector of a subset of models evaluation method computed before
  build.clamping.mask = T, # if TRUE, a clamping mask will be saved on hard drive different
  output.format = '.img') # the format of the GIS files: .RData, .grd or .img

# summary of created object
myBiomodProj
```



###################################################
# Projection to future conditions


```{r}
# Load environmental variables for 2020
layers2020<-stack(list.files(path=path2020,pattern='asc',full.names=TRUE))
# Descriptive statistics for environmental layers
summary(layers2020)
```


# Select variables to include in the model
```{r}
names(layers2020)
egv2020<-layers2020[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]

```


# Plot environmental variables on geographical space
```{r}
plot(egv2020)
```


```{r}
myBiomodProj2020 <- BIOMOD_Projection(
  modeling.output = myBiomodModelOut,
  new.env = egv2020,
  proj.name = 'future2020',
  selected.models = 'all',
  binary.meth = 'ROC',
  build.clamping.mask = T,
  output.format = '.img')

# print summary
myBiomodProj2020
```



# files created on hard drive
```{r}
list.files(path2models2020)
```



# Load environmental variables for 2050
```{r}
layers2050<-stack(list.files(path=path2050,pattern='asc',full.names=TRUE))
# Descriptive statistics for environmental layers
summary(layers2050)
```


# Select variables to include in the model
```{r}
names(layers2050)
egv2050<-layers2050[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
```


# Plot environmental variables on geographical space
```{r}
plot(egv2050)
```


```{r}
myBiomodProj2050 <- BIOMOD_Projection(
  modeling.output = myBiomodModelOut,
  new.env = egv2050,
  proj.name = 'future2050',
  selected.models = 'all',
  binary.meth = 'ROC',
  build.clamping.mask = T,
  output.format = '.img')

```


# print summary
```{r}
myBiomodProj2050
```



# files created on hard drive
```{r}

list.files(path2models2050)
```


# Load environmental variables for 2080
```{r}
layers2080<-stack(list.files(path=path2080,pattern='asc',full.names=TRUE))
# Descriptive statistics for environmental layers
summary(layers2080)
```


# Select variables to include in the model
```{r}
names(layers2080)
egv2080<-layers2080[[c("bio_03","bio_07","bio_08","bio_10","bio_16","bio_18")]]
```


# Plot environmental variables on geographical space
```{r}
plot(egv2080)
```


```{r}
myBiomodProj2080 <- BIOMOD_Projection(
  modeling.output = myBiomodModelOut,
  new.env = egv2080,
  proj.name = 'future2080',
  selected.models = 'all',
  binary.meth = 'ROC',
  build.clamping.mask = T,
  output.format = '.img')

# print summary
myBiomodProj2080

```


# files created on hard drive
```{r}
list.files(path2models2080)
```



# Plot Bioclim current and future models
```{r}
plot(myBiomodProj)
plot(myBiomodProj2020)
plot(myBiomodProj2050)
plot(myBiomodProj2080)
```


RF forest worked better , this one selected areas where species cant exist

```{r}
myRespPlot2D <- response.plot2(models  = myRF,
                               Data = get_formal_data(myBiomodModelOut,'expl.var'), 
                               show.variables= get_formal_data(myBiomodModelOut,'expl.var.names'),
                               do.bivariate = FALSE,
                               fixed.var.metric = 'median',
                               col = c("blue", "red"),
                               legend = TRUE,
                               data_species = get_formal_data(myBiomodModelOut,'resp.var'))
```

