---
title: "ENM Day 5: Ecospat"
output: 
  html_notebook:
    number_sections: true
    theme: cerulean
    toc: true
---
Prsense only model

# Niche Quantification with Ordination techniques

load ecospat package
```{r, results='hide'}
library(ecospat)
library(raster)
library(here)
```

```{r}
# path to layers from area 1
path2area1<-here("layers","pyrenees","area_1")
# path to layers from area 2
path2area2<-here("layers","pyrenees","area_2")
# path to species 1
path2sp1<-here("layers","pyrenees","species1.csv")
# path to species 2
path2sp2<-here("layers","pyrenees","species2.csv")
# path to plot 1
path2plot1<-here("results","ecospat","plot_pca_env_1.jpg")
# path to plot 2
path2plot2<-here("results","ecospat","plot_pca_env_2.jpg")
```


Load environmental layers

not sensitive to correlation in env variables
usulaly you do teh models and then ecospat to compar ethe things
```{r}
egv1<-stack(list.files(path=path2area1, pattern='asc',full.names=TRUE ))
# Descriptive statistics for environmental layers
summary(egv1)
```

Plot variables
```{r}
plot(egv1)
```

## Load environmental layers
```{r}
egv2<-stack(list.files(path=path2area2,pattern='asc',full.names=TRUE ))
# Descriptive statistics for environmental layers
summary(egv2)
```


# Plot variables
```{r}

plot(egv2)
```


# Sample Environmental Variables

Ecospat does not work with rasters 
- for species 1
```{r}
clim1<-data.frame(rasterToPoints(egv1))
head(clim1)
```

- for species 2
```{r}
clim2<-data.frame(rasterToPoints(egv2))
head(clim2)
```

- global climate for both ranges

combine them
```{r}
clim12<-rbind(clim1,clim2)
head(clim12)
```

## loading occurrence sites for the species (column names should be x,y)

- species 1
```{r}
occ.sp1<-na.exclude(read.delim(path2sp1,h=T,sep=","))
```
  export in qgic save as chose comma separated file or load directly shape file 
  
  check species 1
```{r}

head(occ.sp1)
str(occ.sp1)
```
OCC is species ID - must be numerical 

  Plot species 1
```{r}
plot(occ.sp1$x,occ.sp1$y)
```


# species 2
```{r}
occ.sp2<-na.exclude(read.delim(path2sp2,h=T,sep=","))
# check species 2
head(occ.sp2)
str(occ.sp2)
```



# Plot species 2
```{r}
plot(occ.sp2$x,occ.sp2$y)
```




## Add environmental values to the species dataframe 1 with ecospat.sample.envar
ARGUMENTS
dfsp: species dataframe with x, y and optional other variables
colspxy: the range of columns for x and y in dfsp
colspkept: the columns of dfsp that should be kept in the final dataframe (by default: xy )
dfvar: environmental dataframe with x, y and environmental variables
colvarxy: the range of columns for x and y in dfvar
colvar: the range of enviromental variables columns in dfvar. (by default: all exept xy )
resolution: distance between x,y of species and environmental datafreme after which values shouldn't be added 
```{r}

# resolution should be the resolution of the climate data grid
occ.sp1<-na.exclude(ecospat.sample.envar(dfsp=occ.sp1, # occurance dataset
                                         colspxy=2:3, # coordinates
                                         colspkept=1:3, # col to maintain: species ID
                                         dfvar=clim1, 
                                         colvarxy=1:2,#coordinates
                                         colvar=3:5,
                                         resolution=0.008333334))
```


## Add environmental values to the species dataframe 2 with ecospat.sample.envar
 resolution should be the resolution of the climate data grid
```{r}
occ.sp2<-na.exclude(ecospat.sample.envar(dfsp=occ.sp2,
                                         colspxy=2:3,
                                         colspkept=1:3,
                                         dfvar=clim2,
                                         colvarxy=1:2,
                                         colvar=3:5,
                                         resolution=0.008333334))
```



## join both species dataframes
```{r}
occ.sp<-rbind(occ.sp1,occ.sp2)
# check species
head(occ.sp)
str(occ.sp)
```


# Plot species
```{r}
plot(occ.sp$x,occ.sp$y)
```


# list of species
```{r}
sp.list<-levels(occ.sp[,1])
sp.list
```



# selection of variables to include in the analyses
# number of variables in clim12
```{r}
Xvar<-c(3:5)
nvar<-length(Xvar)
```


#number of interation for the tests of equivalency and similarity
```{r}
iterations<-100
#resolution of the gridding of the climate space
R=100 
```
R is 100 minimum, otherwise the kernel density plot will vbe pixelized


## PCA-ENVIRONMENT 

Jion species with a climate ,
we need to define 
```{r}
data<-rbind(occ.sp[,Xvar+1],clim12[,Xvar])
#dataset for the analysis,includes all the sites of the study area + the occurences for all the species
head(data)

```


# vector of weight, 0 for the occurences, 1 for the sites of the study area
```{r}
w<-c(rep(0,nrow(occ.sp)),rep(1,nrow(clim12)))
head(w)
```


# the pca is calibrated on all the sites of the study area
 occurences are not used for the calibration, but their scores are calculated
```{r}
pca.cal <-dudi.pca(data, row.w = w, center = T, scale = T, scannf = F, nf = 2)
```



# selection of species ######
```{r}
sp.list
sp.combn<-combn(1:2,2)
```
1:10 , depending on how many species (n=10)

 indicate where is the species field in occ.sp
```{r}
for(i in 1:ncol(sp.combn)) {
  row.sp1<-which(occ.sp[,1] == sp.list[sp.combn[1,i]]) # rows in data corresponding to sp1
  row.sp2<-which(occ.sp[,1] == sp.list[sp.combn[2,i]]) # rows in data corresponding to sp2
  name.sp1<-sp.list[sp.combn[1,i]]
  name.sp2<-sp.list[sp.combn[2,i]]
  # predict the scores on the axes
  scores.clim<- pca.cal$li[(nrow(occ.sp)+1):nrow(data),]
  #scores for global climate
  scores.sp1<- pca.cal$li[row.sp1,]
  #scores for sp1
  scores.sp2<- pca.cal$li[row.sp2,]
  #scores for sp2
}
```

## Dynamic Occurrence Densities Grid ecospat.grid.clim.dyn
```{r}
z1.dyn<-ecospat.grid.clim.dyn (scores.clim, scores.clim, scores.sp1, R=100)
z2.dyn<-ecospat.grid.clim.dyn (scores.clim, scores.clim, scores.sp2, R=100)
```
kernel


## Niche Equivalency Test ecospat.niche.equivalency.test
```{r}
a.dyn<-ecospat.niche.equivalency.test(z1=z1.dyn , z2=z2.dyn, rep=100)
```



## Niche Similarity Test ecospat.niche.similarity.test
```{r}
b.dyn<-ecospat.niche.similarity.test(z1=z1.dyn , z2=z2.dyn, rep=100)
b2.dyn<-ecospat.niche.similarity.test(z1=z2.dyn , z2=z1.dyn, rep=100)
```

1st is 1 to 2, 
2nd is 2 to 1

### FIRST PLOT

```{r}
# Save the first plot
#jpeg(path2plot1, width = 1500, height = 1500, units = "px", pointsize = 30,quality =100)
# Make a 2x2 plot
par(mfrow = c(2, 2))

# Plot Species Density ecospat.plot.niche for the first species
z1<- ecospat.grid.clim.dyn(scores.clim,scores.clim,scores.sp1,R)
# The plot shows the occupancy of species 1 (grey gradient) and environmental availability in the study area (solid line shows 100% of available climates; dashed line shows 50% most frequent available climates)
ecospat.plot.niche (z1, title="sp1", name.axis1="PC1", name.axis2="PC2",cor=F)

# Plot Species Density ecospat.plot.niche for the second species
z2<- ecospat.grid.clim.dyn(scores.clim,scores.clim,scores.sp2,R)
# The plot shows the occupancy of species 2 (grey gradient) and environmental availability in the study area (solid line shows 100% of available climates; dashed line shows 50% most frequent available climates)
ecospat.plot.niche (z2, title="sp2", name.axis1="PC1", name.axis2="PC2",cor=F)

# Niche Categories and Species Density ecospat.plot.niche.dyn
# Green pixels indicate the unfilled niche (native only), blue pixels the stable niche (common between native and invasive) and red pixels the expansion of the niche (invasive only)
ecospat.plot.niche.dyn(z1.dyn, z2.dyn, title="Niche Categories and Species Density",quant=0.75)

# Draw Centroid Arrows
ecospat.shift.centroids (scores.sp1, scores.sp2, clim1,clim2)
# The arrow indicates the change in the centroid of the niche



```
the following  line of code 
$dynamic.index.w
expansion( only sp1) stability (share) unfilling (only sp2)
```{r}
test<-ecospat.niche.dyn.index (z1.dyn, z2.dyn, intersection=NA)
test$dynamic.index.w
```


### SECOND PLOT
```{r}
# Save the first plot
#jpeg(path2plot2, width = 1500, height = 1500, units = "px", pointsize = 30,quality =100)
# Make a 2x2 plot 
par(mfrow = c(2, 2))
# Plot Variables Contribution ecospat.plot.contrib
ecospat.plot.contrib(contrib=pca.cal$co, eigen=pca.cal$eig)


# Plot Overlap Test ecospat.plot.overlap.test Equivalency
# The plot indicates that the observed overlap is lower than 95% of simulated overlaps. The assumption of niche equivalency is thus rejected
ecospat.plot.overlap.test(a.dyn,"D","Equivalency")

# Similarity 2->1
# The plot indicates that the observed overlap is within 95% of simulated overlaps. The assumption of niche similarity thus cannot be rejected
ecospat.plot.overlap.test(b.dyn,"D","Similarity 1->2")

# Similarity 1->2
# The plot indicates that the observed overlap is within 95% of simulated overlaps. The assumption of niche similarity thus cannot be rejected
ecospat.plot.overlap.test(b2.dyn,"D","Similarity 2->1")

```
similarity: yes, they are different, but not far from null values (p=1), test is not significanto is close to 0.4
similarity: species are different because they are different study areas (p=1)
no stat significant: because it is teh same region
```{r}
# The correlation circle indicate the contribution of original predictors to the PCA axes
ecospat.niche.overlap (z1=z1.dyn, z2=z2.dyn, cor=TRUE)

```

